---
title: "Úvodní hodina"
subtitle: "Zpracovávání výzkumných dat pomocí jazyka R"
title-slide-attributes:
  data-background-color: rgb(233, 84, 32)
format:
  revealjs:
    self-contained: true
    theme: default
    scrollable: true
    footer: "Charleston | [Lucie Hošková](https://orcid.org/0009-0001-9695-2342)"
engine: knitr
filters:
  - webr
execute:
  echo: true
  warning: false
  cache: true
---

## Programme

::: incremental
- short introduction and organization window aka Where Can You Find The Coffee
- R basics 101 review
- so what type of data am I actually working with and why do I need to know?
- ukládání dat do proměnných
- operátory
- etiketa psaní kódu
- working directory
- RProjekty, .RData
- funkce v R a jejich syntax
:::

## Organization

::: incremental
- four 45 minutes long blocks with breaks inbetween

- combination of code along, presentation and individual work

- focused on those who already know the basics of operating witR bash R (and RStudio)
:::

## R basics recap - what do you know?

**What does the symbol \$ do?**

\$ selects values of a particular column from the dataset

dataset\$columnName

**Which deliminator is used in R?**

Decimal point; however, that does not mean that you cannot use datasets which use comma.

Which function can we use to load datasets from .txt or .csv files into R?

`read.table()`

For txt: `read.delim()`

For CSV: `read.csv()`

What is a factor?

## Types of data

why it is important to know the difference

qualitaitve data diskrétní - nominal data - ordinal data

quantitative data spojitá - interval data - pokracuje to jak je treba, -25 stupnu je porad hodnota, nema to neco jako jasnou nulu - nelze porovnavat - teplota v celsius nemuze byt napriklad dvakrat tak vetsi, protoze... dvakrat tak vetsi od ceho? - 40C neni 2x vetsi nez 20C, protoze nula neni zacatek od ktereho se to pocita - ratio - jasna nula; nelze mit zapornou vysku nebo treba prijem, taky nelze jit do minusu se stupni kelvina

is it possible to change those types? yes fo example dividing height into categories

## Loading the data

```{r}
#| echo: false
#| include: false

dataNewborns <- read.delim("newborns.txt")
dataNewborns$edu.M <- factor(dataNewborns$edu.M, labels = c("ZS", "SS", "SSm", "VS"))
dataNewborns$sex.C <- as.factor(dataNewborns$sex.C)
```

During this workshop, we will be working with many types of datasets, most of which are saved in various packages. The first one that we will be however using is the dataset in the file [`newborns.txt`](https://luc-hoskova.github.io/R-individualni_kurzy/Charleston-visualization/newborns.txt).

To be able to create visualizations based on the information within, we first have to **load it into our instance of R**.

::: callout-important
## Always check your data!

Just loading the data is never enough; always check the data to make sure that **each variable is the correct data type**; for example, the categorical data are set as factors, not just character or numeric values.
:::

------------------------------------------------------------------------

what function will we use? how will we do it?

what types are the different variables? what function to use changing to factors - if its categories which use numbers for names it will take it as numeric and for example barlots will go haywire

changing to factors

## Example of base R

načtení dat

# why you should use ggplot2 instead

## Barplot

for categorical data or rather - numeric values of the categories categories on x the numeric values on y it is possible to do how much is of each category like with histogram aka count

https://lucie-hoskova.quarto.pub/r_pro_analyzu_vedeckych_dat/letni_skola.html#/barplot

## point

```{r}
#| label: fig-airquality
#| fig-cap: "Temperature and ozone level."
#| warning: false

library(ggplot2)

ggplot(airquality, aes(Temp, Ozone)) + 
  geom_point() + 
  geom_smooth(method = "loess")
```

## 3D graphs

library(rgl)

x \<- rnorm(100) y \<- rnorm(100) z \<- rnorm(100)

plot3d(x, y, z, col = "blue", pch = 20, size = 2, xlab = "X", ylab = "Y", zlab = "Z") #v prezentaci se nezobrazi, musi byt spusten v RStudiu
